### 2025-06-27 paper/quarto_aaai/paper.tex
Adjust introduction

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy \citep{goodfellow2016deep}. Modern advances in computing have
made it possible to provide such models with ever-growing degrees of
freedom to achieve that task, which frequently allows them to outperform
traditionally more parsimonious models. Unfortunately, in doing so,
models learn increasingly complex and highly sensitive representations
that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, \citet{szegedy2013intriguing} showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
\citep{goodfellow2014explaining}---truly robust deep learning remains
unattainable even for models that are considered ``shallow'' by today's
standards \citep{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \citep{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so-called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\citep{frankle2018lottery}. Similarly, \citet{zhang2021understanding}
showed that state-of-the-art neural networks are expressive enough to
fit randomly labeled data. Thus, looking at the predictive performance
alone, the solutions may seem to provide compelling explanations for the
data, when in fact they are based on purely associative, semantically
meaningless patterns. This poses two challenges. Firstly, there is no
dependable way to verify if representations correspond to meaningful,
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models (see,
e.g., \citep{oneil2016weapons}). Effective XAI tools can aid us in
monitoring models and providing recourse to individuals to turn negative
outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
``application accepted''). In line with this, our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\citet{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Def. \ref{def-explainability}). Instead, the choice has generally been
to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are in fact
compatible with plausible explanations \citep{altmeyer2024faithful}.
Fortunately, recent findings indicate that improved explanatory capacity
can arise as a consequence of regularization techniques aimed at other
training objectives such as generative capacity, generalization, or
robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
novel training regime explicitly geared towards improving the
explainability of models. In high-level terms, we define this concept as
as the extent to which valid explanations derived for an opaque model
are also deemed plausible with respect to the underlying data and the
global actionability constraints. To the best of our knowledge, our
framework represents the first attempt to address this challenge by
employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we demonstrate that CT substantially improves
explainability and positively contributes to the robustness of trained
models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss the challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

@@

\begin{table}

@@

\end{table}%

>>>>

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules hand-crafted by humans, modern machine learning (ML)
models are tasked with learning representations directly from the data,
guided by narrow objectives such as predictive accuracy
\citep{goodfellow2016deep}. Advances in computing have made it possible
to provide these models with ever-growing degrees of freedom to achieve
this task, which often allows them to outperform traditionally
parsimonious models. Unfortunately, in doing so, models learn
increasingly complex, sensitive representations that humans can no
longer easily interpret.

The trend towards complexity for the sake of performance has come under
scrutiny in recent years. At the very cusp of the deep learning (DL)
revolution, \citet{szegedy2013intriguing} showed that artificial neural
networks (ANN) are susceptible to adversarial examples (AEs): perturbed
versions of data instances that yield vastly different model predictions
despite being semantically indistinguishable from their factual
counterparts. Some partial mitigation strategies have been
proposed---most notably \textbf{adversarial training}
\citep{goodfellow2014explaining}---but truly robust deep learning
remains unattainable even for models that are considered ``shallow'' by
today's standards \citep{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \citep{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} As one example, research on the
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% without losing predictive performance
\citep{frankle2018lottery}. Thus, looking at the predictive performance
alone, found solutions may seem to provide compelling explanations for
the data, when in fact they are based on purely associative and
semantically meaningless patterns. This poses two related challenges.
Firstly, there is no dependable way to verify if learned representations
correspond to meaningful, plausible explanations. Secondly, even if we
resolve this challenge, it remains undecided how to ensure that machine
learning models can \emph{only} learn valuable explanations.

The first challenge has attracted an abundance of work on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations, aiming to mitigate scenarios in which
practitioners deploy opaque models and have to blindly rely on their
predictions. On many occasions, this has happened in practice, causing
harms to people who were adversely and unfairly affected by automated
decision-making (ADM) systems involving opaque models (see, e.g.,
\citet{oneil2016weapons}). Effective XAI tools can also aid in
monitoring models and providing recourse, empowering people to turn
negative outcomes (e.g., ``loan application rejected'') into positive
ones (e.g., ``loan application accepted''). In line with this, our work
builds upon \textbf{counterfactual explanations} (CE) proposed by
\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output.

To our surprise, the second challenge has not yet attracted major
research interest. In particular, there has been no concerted effort
towards improving the degree to which learned representations correspond
to explanations that are \textbf{interpretable} and deemed
\textbf{plausible} by humans, which we simply term ``explainability'' in
this manuscript (see Def. \ref{def-explainability}). Instead, the choice
has generally been to improve the ability of XAI tools to identify the
subset of explanations that are both plausible and valid for any given
model, independent of whether these explanations are compatible with the
learned representations \citep{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explainability can arise as a
consequence of regularization techniques aimed at other training
objectives such as generative capacity, generalization, or robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
Our work consolidates these findings within a single framework.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explainability
of models. In high-level terms, we define this concept as as the extent
to which valid explanations derived for an opaque model are also deemed
plausible with respect to the underlying data and the global
actionability constraints. To our knowledge, our framework represents
the first attempt to address this challenge by employing counterfactual
explanations already during the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we empirically demonstrate that CT substantially
improves explainability and positively contributes to the robustness of
trained models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss open challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

@@

\begin{table*}

@@

\end{table*}%

### 2025-06-27 paper/sections/intro.qmd
Adjust introduction

Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which frequently allows them to outperform traditionally more parsimonious models. Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are expressive enough to fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two challenges. Firstly, there is no dependable way to verify if representations correspond to meaningful, plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@mcgregor2021preventing; @oneil2016weapons]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are in fact compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.

**Specifically, we introduce Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models. In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints. To the best of our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already in the training phase.

 The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance. We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Advances in computing have made it possible to provide these models with ever-growing degrees of freedom to achieve this task, which often allows them to outperform traditionally parsimonious models. Unfortunately, in doing so, models learn increasingly complex, sensitive representations that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under scrutiny in recent years. At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are susceptible to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being semantically indistinguishable from their factual counterparts. Some partial mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---but truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery]. Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations. Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.

The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions. On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., @oneil2016weapons). Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.

To our surprise, the second challenge has not yet attracted major research interest. In particular, there has been no concerted effort towards improving the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans, which we simply term "explainability" in this manuscript (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explainability can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating]. Our work consolidates these findings within a single framework.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models. In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints. To our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already during the training phase.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance. Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.
